---
title: "Answering Business Questions using SQL"
author: 'Kate Blaine'
date: 'June 3rd, 2020'
output: html_notebook
---

In this project, I will analyze data from a record company to answer business questions.

```{r}
library(DBI)
library(RSQLite)
library(ggplot2)
```

First, we create helper functions for running SQL queries.

```{r}
run_query <- function(q){
  conn <- dbConnect(SQLite(), 'chinook.db')
  res <- dbSendQuery(conn, q)
  result <- dbFetch(res)
  dbDisconnect(conn)
  return(result)
}
show_tables <- function(){
  q = "SELECT name, type FROM sqlite_master WHERE type IN ('table', 'view')"
  return(run_query(q))
}
show_tables()

```

The following query retrieves the number of tracks sold for each genre in the US.

```{r}
genre_sales <- '
WITH usa_tracks_sold AS 
  (
  SELECT il.* FROM invoice_line il
      INNER JOIN invoice i on il.invoice_id = i.invoice_id
      INNER JOIN customer c on i.customer_id = c.customer_id
      WHERE c.country = "USA"
  )
SELECT 
  g.name genre, 
  COUNT(uts.invoice_line_id) AS number_sold,
  CAST(COUNT(uts.invoice_line_id) AS Float) / (SELECT COUNT(*) FROM usa_tracks_sold)
   percentage_sold
FROM usa_tracks_sold uts
  INNER JOIN track t on t.track_id = uts.track_id
  INNER JOIN genre g on g.genre_id = t.genre_id
GROUP BY genre
ORDER BY number_sold DESC
LIMIT 10;
'
genre_data <- run_query(genre_sales)
```

We will visualize the results with a bar chart.

```{r}
ggplot(data = genre_data) +
  aes(x = genre, y = percentage_sold) +
  geom_bar(stat = 'identity')

```

Clearly, the most tracks sold belong to the rock genre. Next, we'll analyze employee performance based on purchases of customers belonging to each sales support agent.

```{r}
employee_query <- '
WITH support_sales AS
  (
  SELECT
    i.customer_id,
    c.support_rep_id,
    SUM(i.total) total
  FROM invoice i
  INNER JOIN customer c ON i.customer_id = c.customer_id
  GROUP BY 1, 2
  )
SELECT
  e.first_name || " " || e.last_name employee,
  e.hire_date, 
  SUM(ss.total) total_sales
FROM support_sales ss
INNER JOIN employee e ON e.employee_id = ss.support_rep_id
GROUP BY 1;
'

employee_sales <- run_query(employee_query)
employee_sales
```

We will visualize the results with a bar chart.

```{r}
ggplot(data = employee_sales) + 
  aes(x = employee, y = total_sales) +
  geom_bar(stat = 'identity')
```

According to the bar chart, Jane Peacock has had the most sales. However, of the 3 employees, Jane Peacock was the first to start at the company, followed by Margaret Park and then Steve Johnson. The difference in sales may be affected by the amount of time the employee has worked at the company.

Next, we'll analyze sales by country. 
```{r}
country_query <- '
WITH country_or_other AS
  (
  SELECT 
    CASE
      WHEN(
        SELECT COUNT(*) FROM customer
        WHERE country = c.country
      ) = 1 THEN "Other"
      ELSE c.country
    END AS country,
    c.customer_id,
    il.*
  FROM invoice_line il
  INNER JOIN invoice i ON i.invoice_id = il.invoice_id
  INNER JOIN customer c ON c.customer_id = i.customer_id
  )
SELECT 
  country,
  customers,
  total_sales,
  average_order,
  customer_lifetime_value
FROM
  (
  SELECT
    country,
    COUNT(DISTINCT customer_id) customers,
    SUM(unit_price) total_sales,
    SUM(unit_price)/COUNT(DISTINCT customer_id) customer_lifetime_value,
    SUM(unit_price)/COUNT(DISTINCT invoice_id) average_order,
    CASE
      WHEN country = "Other" THEN 1
      ELSE 0
    END AS sort
    FROM country_or_other
    GROUP BY country
    ORDER BY sort ASC, total_sales DESC
  );
  '
country_data <- run_query(country_query)
country_data
```

Let's visualize this data using 3 bar charts: one depicting total sales, one for total customers, and one for average customer orders.

```{r}
ggplot(data = country_data) +
  aes(x = country, y = total_sales, fill = country) +
  geom_bar(stat = 'identity') +
  labs(title = 'Total sales by country', x = 'Country', y = 'Total sales')

ggplot(data = country_data) +
  aes(x = country, y = customers, fill = country) +
  geom_bar(stat = 'identity') + 
  labs(title = 'Number of customers by country', x = 'Country', y = 'Customers')

ggplot(data = country_data) +
  aes(x = country, y = customer_lifetime_value, fill = country) +
  geom_bar(stat = 'identity') +
  labs(title = 'Customer lifetime value by country', x = 'Country', y = 'Customer lifetime value')
```

Based on these bar charts, we can see that the United States has the most sales and customers, but the customers don't necessarily have very large orders. On the other hand, the Czech Republic has a lower number of customers, but those customers are making relatively large orders. There may be an opportunity for expansion in the Czech Republic to bring in more customers.

Finally, we're going to explore whether customers are more likely to purchase a whole album or individual tracks from an album.
```{r}
album_query <- '
WITH invoice_first_track AS
  (
  SELECT
    il.invoice_id invoice_id,
    MIN(il.track_id) first_track_id
  FROM invoice_line il
  GROUP BY 1
  )
SELECT
  album_purchase,
  COUNT(invoice_id) number_of_invoices,
  CAST(COUNT(invoice_id) AS Float) / (SELECT COUNT(*) FROM invoice) percent
FROM (
SELECT
  ifs.*,
  CASE
    WHEN
      (
      SELECT t.track_id FROM track t
      WHERE t.album_id = (SELECT t2.album_id FROM track t2 WHERE t2.track_id = ifs.first_track_id)
      
      EXCEPT
      
      SELECT il2.track_id FROM invoice_line il2
      WHERE il2.invoice_id = ifs.invoice_id
      ) IS NULL
    AND
      (
      SELECT il2.track_id FROM invoice_line il2
      WHERE il2.invoice_id = ifs.invoice_id
      
      EXCEPT
      
      SELECT t.track_id FROM track t
      WHERE t.album_id = (SELECT t2.album_id FROM track t2 WHERE t2.track_id = ifs.first_track_id)
      ) IS NULL
    THEN "Yes"
    ELSE "No"
  END AS album_purchase
FROM invoice_first_track ifs
      )
GROUP BY album_purchase;
'
album_data <- run_query(album_query)
album_data
```

According to the results, full album purchases account for about 18% of all purchases. Therefore, the company should continue buying full albums from record companies.
